\section{Environmental Factors in Plant Growth (approx. 3 Pages)}

\subsection{Importance of Soil Moisture, Temperature, Light, and Humidity}
\begin{itemize}
    \item \textbf{Soil Moisture:}
    \begin{itemize}
        \item Required for nutrient transport and root metabolism.
        \item Direct influence on oxygen availability in the root zone.
        \item Too much water leads to oxygen deficiency and root rot.
        \item Too little water causes dehydration and reduced nutrient uptake.
    \end{itemize}

    \item \textbf{Light:}
    \begin{itemize}
        \item Primary energy source for photosynthesis.
        \item Intensity, duration, and spectrum influence plant growth.
        \item Difference between shade plants and sun plants.
        \item Indoor environments often provide insufficient usable light.
    \end{itemize}

    \item \textbf{Temperature and Humidity:}
    \begin{itemize}
        \item Temperature affects metabolic activity and growth speed.
        \item Humidity influences transpiration and water loss through leaves.
        \item Extreme values cause plant stress.
        \item Interaction between temperature and humidity (dry air increases water loss).
    \end{itemize}
\end{itemize}

\subsection{Effects of Environmental Factors on Plant Growth}
\begin{itemize}
    \item \textbf{Deficiency:}
    \begin{itemize}
        \item Wilting caused by insufficient water uptake.
        \item Slow or stunted growth due to lack of energy or nutrients.
        \item Etiolation (elongated, weak growth) caused by insufficient light.
        \item Increased vulnerability to pests and diseases.
    \end{itemize}

    \item \textbf{Excess:}
    \begin{itemize}
        \item Mold and fungal growth caused by excessive moisture.
        \item Root suffocation due to waterlogged soil.
        \item Leaf burn or stress caused by excessive light or heat.
        \item Long-term damage through continuous stress conditions.
    \end{itemize}
\end{itemize}

\subsection{Necessity of Precise Measurements}
\begin{itemize}
    \item Visual or tactile assessment of soil moisture is unreliable.
    \item Soil surface conditions often differ from moisture levels in the root zone.
    \item Human judgment of light intensity is inaccurate due to automatic eye adaptation.
    \item Environmental conditions change over time and cannot be captured by occasional checks.
    \item Objective sensor measurements provide consistent and comparable data.
    \item Continuous monitoring enables trend detection instead of single snapshots.
\end{itemize}

\subsection*{Possible Sources for Environmental Factors in Plant Growth(einfach mal reingegeben, wird sich später genauer angeschaut)}
\begin{itemize}
    \item Overwatering and underwatering symptoms in houseplants:  
    \url{https://localbotanist.com.au/news/deciphering-plant-signals-the-telltale-signs-of-overwatering-vs-underwatering/}

    \item General reasons for plant failure and care mistakes:  
    \url{https://www.squarely-copenhagen.com/blogs/news/7-common-reasons-why-your-plants-are-dying-how-you-can-save-them}

    \item Soil moisture influence on nutrient availability:  
    \url{https://www.internationalscholarsjournals.com/articles/soil-moisture-influence-on-nutrient-availability-in-the-soil.pdf}

    \item Maximum plant uptake of water, nutrients, and oxygen:  
    \url{https://pmc.ncbi.nlm.nih.gov/articles/PMC5387063/}

    \item Impact of light intensity and spectrum on plant growth:  
    \url{https://pmc.ncbi.nlm.nih.gov/articles/PMC8538153/}

    \item Photosynthetically active radiation and usable light for plants:  
    \url{https://www.2hraquarist.com/blogs/light-3pillars/light-wavelength-indepth}

    \item Plant transpiration and water loss mechanisms:  
    \url{https://emeraldharvest.co/understanding-plant-transpiration/}

    \item Effects of heat stress on transpiration and plant physiology:  
    \url{https://pmc.ncbi.nlm.nih.gov/articles/PMC9892838/}

    \item Etiolation caused by insufficient light:  
    \url{https://www.sciencedirect.com/topics/biochemistry-genetics-and-molecular-biology/etiolation}

    \item Abiotic stress caused by environmental deficiency and excess:  
    \url{https://www.frontiersin.org/journals/plant-science/articles/10.3389/fpls.2025.1545912/full}

    \item Low-cost soil moisture monitoring and measurement methods:  
    \url{https://attra.ncat.org/publication/soil-moisture-monitoring-low-cost-tools-and-methods/}
\end{itemize}
