\section{Data Acquisition and Processing (approx. 4 Pages)}

\subsection{Data Acquisition with Microcontrollers}
\begin{itemize}
    \item \textbf{Controller Selection:}
    \begin{itemize}
        \item ESP32 selected due to built-in Wi-Fi and Bluetooth.
        \item Well suited for IoT projects and battery-powered systems.
        \item Supports deep sleep modes to reduce power consumption.
    \end{itemize}

    \item \textbf{Sensor Inputs and Interfaces:}
    \begin{itemize}
        \item Soil moisture sensors connected via analog inputs (ADC).
        \item Environmental sensors connected using the I\textsuperscript{2}C interface (e.g. BME280).
        \item Using separate interfaces improves measurement reliability.
    \end{itemize}

    \item \textbf{Sampling Strategy:}
    \begin{itemize}
        \item Sensors are read at regular time intervals.
        \item Sampling rate chosen to balance detail and battery life.
        \item Timer-based wake-up used between measurements.
    \end{itemize}
\end{itemize}

\subsection{Signal Processing and Data Transmission}
\begin{itemize}
    \item \textbf{Raw Sensor Data:}
    \begin{itemize}
        \item Sensor readings can vary due to noise and changing conditions.
        \item Single values may not represent actual plant conditions.
    \end{itemize}

    \item \textbf{Data Processing:}
    \begin{itemize}
        \item Multiple measurements taken during each cycle.
        \item Average values calculated to smooth fluctuations.
        \item Data processing performed directly on the microcontroller.
    \end{itemize}

    \item \textbf{Data Transmission:}
    \begin{itemize}
        \item Sensor data formatted as JSON.
        \item Lightweight data structure suitable for small devices.
        \item Data sent to the backend using MQTT or HTTP.
    \end{itemize}
\end{itemize}

\subsection{Fundamentals of Data Cleaning}
\begin{itemize}
    \item Removal of invalid values (e.g. negative humidity).
    \item Handling of missing or incomplete data.
    \item Filtering of outliers caused by temporary sensor issues.
    \item Ensuring data consistency before further processing.
\end{itemize}

\subsection*{Possible Sources for Data Acquisition and Processing}

\begin{itemize}
    \item ESP32 deep sleep modes and power consumption:  
    \url{https://deepbluembedded.com/esp32-sleep-modes-power-consumption/}

    \item Practical ESP32 deep sleep tutorial:  
    \url{https://www.instructables.com/ESP32-Deep-Sleep-Tutorial/}

    \item Introduction to reading and processing sensor data on microcontrollers:  
    \url{https://www.freecodecamp.org/news/connect-read-process-sensor-data-on-microcontrollers-for-beginners/}
    
    \item JSON data formatting for IoT applications:  
    \url{https://www.engineersgarage.com/arduino-json-http-requests-iot/}

    \item Efficient JSON serialization for embedded systems:  
    \url{https://arduinojson.org}

    \item Comparison of MQTT and HTTP for IoT data transmission:  
    \url{https://www.emqx.com/en/blog/mqtt-vs-http}

    \item Time-series databases and InfluxDB fundamentals:  
    \url{https://www.influxdata.com/time-series-database/}

    \item Importance of data quality in time-series databases:  
    \url{https://www.influxdata.com/blog/bad-data-costs-influxdb/}

    \item Moving average filtering for sensor noise reduction:  
    \url{https://codemonk.io/blog/moving-average-filter/}

    \item Handling missing values in environmental sensor data:  
    \url{https://pmc.ncbi.nlm.nih.gov/articles/PMC11053546/}
\end{itemize}

