\section{System Implementation and Validation (approx. 5--6 Pages)}

\subsection{Hardware Realization and Setup}
\begin{itemize}
    \item \textbf{Power Supply Concept:}
    \begin{itemize}
        \item Battery-powered operation enables flexible placement near plants.
        \item USB power used during development, testing, and debugging.
        \item Trade-off between mobility and long-term stability.
    \end{itemize}

    \item \textbf{Sensor Placement:}
    \begin{itemize}
        \item Soil moisture sensor placed in the root zone, not at the surface.
        \item Avoids misleading surface measurements.
        \item Placement depth adapted to typical indoor plant pots.
    \end{itemize}

    \item \textbf{Enclosure Design:}
    \begin{itemize}
        \item Custom 3D-printed housing for Plant Up!.
        \item Protection against water splashes during watering.
        \item Ventilation openings for air sensors.
        \item Compact form factor suitable for indoor environments.
    \end{itemize}
\end{itemize}

\subsection{Software Architecture and Firmware}
\begin{itemize}
    \item \textbf{Overall System Flow:}
    \begin{itemize}
        \item Wake from deep sleep.
        \item Read sensor values.
        \item Process and prepare data.
        \item Transmit data to backend.
        \item Return to deep sleep.
    \end{itemize}

    \item \textbf{Error Handling and Robustness:}
    \begin{itemize}
        \item Handling of missing or failed sensor readings.
        \item Retry logic for network communication failures.
        \item System stability prioritized over continuous availability.
    \end{itemize}

    \item \textbf{Energy Management:}
    \begin{itemize}
        \item Deep sleep mode used to minimize power consumption.
        \item Wake-up intervals configurable based on plant needs.
        \item Significant battery life improvement compared to continuous operation.
    \end{itemize}
\end{itemize}

\subsection{Sensor Calibration}
\begin{itemize}
    \item Raw sensor values provided as ADC readings (0--4095).
    \item Raw values are not directly meaningful for users.
    \item Calibration process:
    \begin{itemize}
        \item Reference measurement in dry soil.
        \item Reference measurement in fully watered soil.
        \item Linear mapping of raw values to percentage scale (0--100\%).
    \end{itemize}
    \item Calibration depends on soil type and substrate composition.
    \item Plant Up! provides relative moisture values rather than laboratory-accurate measurements.
\end{itemize}

\subsection{Results Presentation: The ``Plant Up!'' Dashboard}
\begin{itemize}
    \item Responsive web-based dashboard.
    \item Focus on clarity and ease of interpretation.
    \item Current plant status displayed prominently.
    \item Use of simple color indicators (green, yellow, red).
    \item Historical data visualized using line charts.
    \item Trend visualization supports long-term care decisions.
\end{itemize}

\subsection{Validation and Measurement Series Analysis}
\begin{itemize}
    \item Test setup:
    \begin{itemize}
        \item Single indoor plant monitored over a 48-hour period.
        \item Controlled watering event during test phase.
    \end{itemize}

    \item Observed behavior:
    \begin{itemize}
        \item Clear increase in soil moisture after watering.
        \item Gradual decrease over time due to evaporation and uptake.
    \end{itemize}

    \item Comparison between expected and measured behavior.
    \item Validation confirms correct sensor response and data processing.
    \item Limitations:
    \begin{itemize}
        \item Short test duration.
        \item Limited number of plants.
    \end{itemize}
    \item Results considered sufficient to validate basic system functionality.
\end{itemize}

\subsection*{Possible Sources for System Implementation and Validation}

\begin{itemize}
    \item ESP32 deep sleep modes and power consumption:  
    \url{https://www.programmingelectronics.com/esp32-deep-sleep-mode/}

    \item Importance of sensor calibration in embedded systems:  
    \url{https://www.a3p.org/en/thecalibration-of-embedded-sensors/}
\end{itemize}
