%!TEX root = ../../Diplomarbeit.tex
\section{Case Study: Gamification in Plant Up!}
This section details the practical implementation of the theoretical concepts within the Plant Up! project.

\subsection{Overview of Plant Up!}
\begin{itemize}
    \item \textbf{Smart gardening system:} An integrated solution creating a bridge between nature and technology.
    \item \textbf{IoT sensors:} Hardware sensors monitor soil moisture, light, and temperature in real-time.
    \item \textbf{Mobile application:} The user interface for monitoring data and interacting with the gamified elements.
    \item \textbf{Target users:} Hobby gardeners, tech enthusiasts, and people struggling to keep plants alive.
\end{itemize}

\subsection{Gamification Concept in Plant Up!}
The concept applies map theory to implementation:
\begin{itemize}
    \item \textbf{XP for plant care:} Users earn Experience Points (XP) for performing correct care actions.
    \item \textbf{Streaks for daily check-ins:} Encouraging regular monitoring through consecutive daily login rewards.
    \item \textbf{Quests:} Specific challenges such as watering, fertilizing, or monitoring parameters.
    \item \textbf{Titles \& Ranks:} Progression system from 'Seedling' to 'Evergreen' to 'Bloom'.
    \item \textbf{Coins \& Lootboxes:} Virtual currency and random rewards to introduce variability.
    \item \textbf{NPC Guide ('Sprig'):} A character that provides tutorials, hints, and narrative context.
\end{itemize}

\subsection{Mapping Design Patterns to Motivation}
Each gamification element is designed to target specific motivational drivers:
\begin{itemize}
    \item \textbf{Streaks \textrightarrow\ Habit Formation:} Builds daily routines through loss aversion.
    \item \textbf{XP \textrightarrow\ Competence:} Provides a quantifiable metric of skill and dedication.
    \item \textbf{Social Feed \textrightarrow\ Relatedness:} Allows users to share success and feel part of a community.
    \item \textbf{Titles \textrightarrow\ Identity \& Status:} Public symbols of achievement and expertise.
\end{itemize}
